## Subject-level variables
In a proper PK/PD analysis we need to explore the data at multiple
variability levels. What we looked at above is at dosing and sampling
level (one row per dosing or sampling event). `NMdata` provides very
useful functions to extract information at other levels of
variability. Before extracting the subject-level information we will
add an individual exposure measure to the dataset. We will use the
empirical Bayes' estimate of the individual maximum
concentration. This is derived as the maximum prediction across the
sample times - it may be better to simulate the model at a richer time
scale to get better precision. Again, we make use of `data.table` but
feel free to use what you prefer. `findCovs` - like other `NMdata`
functions - will work on `data.frame`'s, `tibble`'s and other
`data.frame`-like classes 
(classes that inherit from `data.frame`).

```{r}
res1 <- NMscanData(system.file("examples/nonmem/xgxr001.lst", package="NMdata"),
                   col.row="ROW",merge.by.row=TRUE,quiet=TRUE,as.fun="data.table")
res1$trtact <- reorder(res1$trtact,res1$DOSE)
## with data.table, create a new column representing ID-level Cmax
res1[,Cmax:=max(IPRED),by=.(ID)]
## findCovs picks the columns that do not vary within cols.id. One row
## per value of cols.id.
res1.id <- findCovs(res1,by="ID")
dim(res1.id)
ggplot(res1.id,aes(WEIGHTB,Cmax/DOSE,colour=trtact))+
    geom_point()+
    labs(x="Bodyweight at baseline (kg)")
```

`by` can be longer than one. Include a study column, a treatment column to
support cross-over, or a model
column if you are comparing model fits (see below how to do this very easily).

If your model includes occasion variability, you probably also want to
look at 

```{r}
## we have no occasion variability in this data
## res1.id.occ <- findCovs(res1,by=c("ID","OCC"))
```

The simplest use of `findCovs` is to get variables that are constant
across the whole dataset:


```{r}
findCovs(res1)
```

Let's take a look at what is in the
`res1.id` generated above. It is a mix of variables that vary at subject level and variables
that are constant across the full dataset. 

```{r}
dim(res1.id)
head(res1.id,2)
```


`findCovs` has a counterpart in `findVars` which finds variables that
do vary within constant values of optional `by` columns. To get only the ones that
vary within the dataset (i.e. they are truely subject-level
variables and not say a study number if we only have one study), we can do 

```{r}
res1.id2 <- findVars(res1.id)
dim(res1.id2)
head(res1.id2,2)
```

`findVars` supports the `by` argument too. So you can use
`findVars(res1,by="ID")` to find variables that are non-constant
within (at least one value of) ID. `by` can be of arbitrary length.

Of course, we most often know what covariates or other subject-level
variables to look at, and we would not search for them after running a
nonmem model. But in the situation where you are looking at someone
else's work or you are doing a meta analysis across models where the
data has been coded slightly differently, these simple tools can be
very useful. Or if you like writing generic wrapper functions, you may
find this very handy. Also, remember that if a variable is returned by
these functions, you know that they fullfill the variability
requirement. We know that variables in `res1.id` above do not vary
within `ID`, meaning that no `ID` has more than one value of the
variable. `NA` counts as any other value, so if a value is returned
for subject in `res1.id`, this subject does not have `NA`'s in `res1`.

