---
title: "Data creation tools"
output:  rmarkdown::html_vignette
Suggests: markdown
VignetteBuilder: knitr
vignette: >
  %\VignetteIndexEntry{Data creation tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{ae}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
 ,fig.width=7)

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

```{r,eval=TRUE,include=FALSE}
## library(devtools)
## load_all("C:/Users/delff/working_copies/NMdata")
```

```{r,setup,include=F}
library(NMdata)
NMdataConf(check.time=FALSE)
## NMdataConf(as.fun="data.table")
library(data.table)
library(ggplot2)
theme_set(theme_bw()+theme(legend.position="bottom"))

```
Built `r Sys.Date()` using NMdata `r packageVersion("NMdata")`.

This vignette is still under development. Please make sure to see
latest version available
[here](https://philipdelff.github.io/NMdata/).

## Introduction
Getting data ready for modeling is a crucial and tedious but often
underestimated task. Mistakes during the process of combining data
sets, defining time variables etc. can lead to difficulties during
modeling, need for revisiting data set creation, and in the worst case
where the errors are not quickly realized, a lot of wasted time
working with an erroneos data set. Avoiding those mistakes by
integrating checks into the data creation process is a must.

Furthermore, Nonmem has quite a number of restrictions on the format
of the input data, and problems with the data set is a very common
reason for Nonmem not to behave as expected. When this happens,
debugging can be time-consuming. `NMdata` includes some simple
functions to prevent these situations.

This vignette uses `data.table` syntax for the little bit of data
manipulation performed. However, you don't need to use data.table _at
all_ to use these tools or any tool in `NMdata`. The reason why the
functions return data.table objects is that the data set is a
`data.table` to begin with

```{r}
pk <- readRDS(file=system.file("examples/data/xgxr2.rds",package="NMdata"))
class(pk)
```

If this was a `data.frame` (and not a `data.table`), the `NMdata` functions would keep returning objects
of class `data.frame`.

## Data assembly


### Compare presence and class of columns across data sets
When stacking (rbind) and merging, it is most often necessary to check
if two or more data sets are compatible for the
operation. `compareCols` by defaults lists the columns where the two
data sets differ.


```{r,include=FALSE}
pk.reduced <- copy(pk)
pk.reduced <- pk.reduced[1:(.N%/%2)]
pk.reduced[,CYCLE:=NULL]
pk.reduced[,AMT:=as.character(AMT)]
```

Just to illustrate the output of `compareCols`, a slightly modified
version of the `pk` dataset has been created. On column (`CYCLE`) has
been removed, and `AMT` has been recoded to character. `compareCols`
tells us about exactly these two differences. So before mergeing or
stacking, we may want to recode `AMT` in one of the datasets to get
the class we need, and decide what to do about the missing `CYCLE` in
one of the datasets (if OK, we fill in `NA`.


```{r,eval=TRUE}
compareCols(pk,pk.reduced)
```

### Rename columns based on contents
The model estimation step is heavily dependent (and in Nonmem almost entirely based) on numeric data values. 
The source data will often contain character variables, i.e. columns with non-numeric data values.

If the column names reflect whether the values are numeric, double-checking can be avoided. `renameByContents` renames columns if a function of their contents returns `TRUE`.


```{r,include=FALSE}
pktmp <- copy(pk)
pktmp[,TRTACT:=NULL]
```

```{r}
pk.renamed <- renameByContents(data=pktmp,fun.test=NMisNumeric,fun.rename = tolower, invert.test = TRUE)
```

`NMisNumeric` is a function that tests if the contents are numeric to `Nonmem`. If say the subject ID is of character class, it can be valid to `Nonmem`. Subject ID `"1039"` will be a numeric in Nonmem, `"1-039"` will not. We invert that, and those that Nonmem cannot interpret as numeric become lowercase. We use `compareCols` to illustrate that three columns were renamed:

```{r,eval=TRUE}
compareCols(pktmp,pk.renamed)
```



### Automated checks of merge results 
Merges are a very common source of data creation bugs. As simple as
they may seem, merges likely leave you with an unexpected number of
rows, some repeated or some omitted. `mergeCheck` provides a merge
which is restricted to the specific kind where the rows that come out
of the merge are the exact same as in one of the existing datasets,
only columns added from the second dataset. This limitation of the
scope of the merge allows for a high degree of automated checks of
consistency of the results.


```{r,include=FALSE}
dt.cov <- pk[,.(ID=unique(ID)[1:10])]
dt.cov[,COV:=sample(1:5,size=10,replace=TRUE)]

dt.cov2 <- pk[,.(ID=unique(ID))]
dt.cov2[,COV:=sample(1:5,size=.N,replace=TRUE)]
dt.cov2 <- dt.cov2[c(1,1:(.N-1))]
```

Say we want to add a covariate from a
`dt.cov`. Imagine we just assume that all subjects are represented with one row
each in `dt.cov`, and so we expect the number of rows to be unchanged from `pk`. But we see that only when remembering `all.x=T`, we get the dimensions we had expected (we do not checked that we get the _right_ rows, only that we get the _right number_ of rows):


```{r}
dim(pk)
## the number of unique values of ID in pk
pk[,uniqueN(ID)]
pk[,range(ID)]
## dt.cov has a covariate for some of the subjects
dt.cov
pk2 <- merge(pk,dt.cov,by="ID")
dim(pk2)
## now as expected
pk3 <- merge(pk,dt.cov,by="ID",all.x=TRUE)
dim(pk3)
```

So when creating `pk2` we lost a lot of rows because we forgot the
`all.x` option. And also as we shall now see that checking the
resulting dimensions is not enough. In the following example, `pk4` has the same number of rows as `pk`, but one row is repeated, another must have disappeared.

```{r}
pk4 <- merge(pk,dt.cov2,by="ID")
dim(pk4)
## we now have twice as many rows for this subject
pk[ID==31,.N]
pk4[ID==31,.N]
```

As illustrated by these simple examples, even the simplest merges have
to be rigorously checked, no exceptions. The function call `mergeCheck(x1,x2)`
ensures that the result has exactly one of each rows from `x1` and
nothing else. We can use `...` to pass additional arguments to merge. Here, we need `all.x` to include rows from `x1` where `ID` is not matched by a row in `dt.cov`.

```{r}
pk2 <- try(mergeCheck(pk,dt.cov,by="ID"))
## now as expected
pk3 <- mergeCheck(pk,dt.cov,by="ID",all.x=TRUE)
dim(pk3)
```

The result always has the same row order as the `x1` argument. 

Another problem that the programmer may not realize during a merge is
that column names are shared across `x1` and `x2` (in addition to columns that are being merged by). This will silently create
colmn names like `col.x` and `col.y` in the output. `mergeCheck` will
by default give a warning if that happens. There is an optional argument to tell mergeCheck how many columns should be added by the merge, and `mergeCheck` will fail if another number of columns are added. This can be useful for programming.

There
is only one difference from the behaviour of the `merge.data.frame`
function syntax, being that the `by` argument must always be supplied to `mergeCheck`.

You may think that this will limit your merges, and that you need merges for inner and outer joins etc. You are exactly right - `mergeCheck` is not intended for those merges and does not support them. When that is said, the kind of merges that are supported by `mergeCheck` are indeed very common. To illustrate how common, all merges in the `NMdata` package are performed with `mergeCheck`.


## Exclusion flags
It is good practice not to discard records from a dataset but to flag
them and omit them in model estimation. When reporting the analysis,
we also need to account for how many data records were discarded due
to which criteria. A couple of functons in `NMdata` help you do this
in a way that is easy to integrate with Nonmem. 

The implementation in `NMdata` is based on sequentially checking
exclusion conditions. This means we can summarize how many records and
subjects were excluded from the analysis due to the different
criteria. The information is represented in one numerical column for
Nonmem, and one (value-to-value corresponding) character column for
the rest of us in the resulting data.

### Assign and count flags
For use in Nonmem, the easiest is that inclusion/exclusion is
determined by a single column in data - we call that column `FLAG`
here, but any column name can be used. `FLAG` obviously draws on
information from other columns such as `TIME`, `DV`, and many others,
depending on your dataset and your way of working.


The function that applies inclusion/excluasion rules is called
`flagsAssign`, and it takes a dataset and a data.frame with rules as
arguments.

```{r,include=F}
pk <- readRDS(file=system.file("examples/data/xgxr2.rds", package="NMdata"))
pk[,`:=`(FLAG=NULL,flag=NULL)]
```
```{r}
dt.flags <- fread(text="FLAG,flag,condition
10,Below LLOQ,EVID==0&BLQ==1
100,Negative time,EVID==0&TIME<0")

pk <- flagsAssign(pk,tab.flags=dt.flags,subset.data="EVID==0")
pk[EVID==1,FLAG:=0]
pk[EVID==1,flag:="Dosing"]
```

Your first question may be why `fread` is used to create a data.table
(like `read.csv` to create a data.frame). The reason is that this way, we
can write the information line by line. If you have ten lines in the
dataset above and do `data.frame(FLAG=...,flag=...,condition=...)` it
gets impossible to read what elements of the three columns that
correspond.

`flagsAssign` applies the conditions sequentially and by increasing
value of `FLAG`. `FLAG=0` means that the observation is included in
the analysis. You can use any expression that can be evaluated within
the data.frame. In this case, `BLQ` has to exist in `pk`.

In `Nonmem`, you can include `IGNORE=(FLAG.NE.0)` in `$DATA` or `$INFILE`.

Again, the omission will be attributed to the first condition
matched. Default is to apply the conditions by the order of decreasing
numerical flag value. There is an argument if you prefer the
opposite. However, what cannot be modified is that 0 is the numerical
value for rows that are not matched by any conditions.


What rows to omit from a dataset can vary from one analysis to
another. Hence, the aim with the chosen design is that the inclusion
criteria can be changed and applied to overwrite an existing
inclusion/exclusion selection. For another analysis we want to include
the observations below LLOQ. We have two options. Either we simply
change the `IGNORE` statement given above to `IGNORE=(FLAG.LT.10)`, or
you create a different exclusion flag for that one. If you prefer to
create a new set of exclusion flags, just use new names for the
numerical and the character flag columns so you don't overwrite the
old ones. See help of `flagsAssign` and `flagsCount` for how.


### Summarize data exclusions
An overview of the number of observations disregarded due to the
different conditions is then obtained using `flagsCount`:

```{r}
tab.count <- flagsCount(data=pk[EVID==0],tab.flags=dt.flags)
print(tab.count)
```

`flagsCount` includes a `file` argument to save the the table right
away.


## Finalize data format and write to file
Once you have your dataset in place, `NMdata` provides a few useful
functions. At this point, ideally you check your data for any
inconsistency. If you miss an inconsistency, you need to wait for Nonmem
to finish if the inconsistency is not fatal for Nonmem's ability to
run the model. Then, you may or may not realize the inconcisteny quickly. `NMdata` is still ways from providing a sufficient set of tests
to run, but already includes some checks that are useful and easy to run.

### Automated ordering of columns
The order of columns in Nonmem is important for two reasons. One is
that a character in a variable read into Nonmem will make the run
fail. The other is that there are restrictions on the number of
variables you can read into Nonmem, depending on the
version. `NMorderColumns` tries to put the used columns first, and
other or maybe even unusable columns in the back of the dataset. It
does so by a mix of recognition of column names and analysis of the
column contents. 

Columns that cannot be converted to numeric are put in the back, while
column bearing standard Nonmem variable names like `ID`, `TIME`, `EVID` etc.
will be pulled up front. You can of course add column names to
prioritize to front (`first`) or back (`last`). See `?NMorderColumns` for
more options.

```{r}
pk <- NMorderColumns(pk)
```

We may want to add `MDV` and rerun `NMorderColumns`.


### Writing data to files
For the final step of writing the dataset, `NMwriteData` is
provided. Most importantly, it writes a csv file with appropriate options for Nonmem to read it as well as possible. It can also write an rds for
R with equal contents (or RData if you prefer), but with the rds including all information
(such as factor levels) which cannot be saved in csv. If you should use `NMscanData` to read Nonmem results,  this information can be used automatically. It also provides a proposal for text to include in the
`$INPUT` and `$DATA` sections of the Nonmem control
streams. 

```{r}
NMwriteData(pk)
```
Notice, `NMwriteData` detected the exclusion flag and suggests to include it in `$DATA`.

If a file name had been provided, the data would have been written,
and the path to the data file would have been included in the message
written back to the user. There are several arguments that will affect
the proposed text for the Nonmem run, see `?NMwriteData`.

### Updating Nonmem control streams to read new data file
I may be the only one, but sometimes during the modeling stage, I want to go back and change or add something to the data creation step. Then once I have written a new data file, my Nonmem `$INPUT` sections no longer match the data file. In `NMwriteData` you can use a `last` argument to get columns pushed towards the back so the Nonmem runs should still work, but maybe you need the column in your nonmem runs, and so you have no way around updating the control streams. And that can be quite a lot of control streams.

`NMdata` has a couple of functions to extract and write sections to Nonmem control streams called `NMreadSection` and `NMwriteSection`. We are not going into detail with what these functions can do, but let's stick to the example above. We can do

```{r,eval=FALSE}
NMwriteSection("run001.mod","INPUT","$INPUT ROW ID TIME EVID CMT AMT DV FLAG STUDY BLQ CYCLE DOSE FLAG2 NOMTIME PART PROFDAY PROFTIME WEIGHTB eff0")
```

But in fact, we can go a step further and take the information straight from NMwriteData

```{r}
text.nm <- NMwriteData(pk)
```

NMwriteData invisibly returns a list of sections (`$INPUT` and `$DATA`). `NMwriteSection` can use these directly. So to write only the `$INPUT` section to `run001.mod`, we do the following. Please notice the single brackets in `text.nm["INPUT"]` which mean that we still send a list to `NMwriteSection`.

```{r,eval=FALSE}
NMwriteSection("run001.mod",list.sections=text.nm["INPUT"])
```

If you run this in a loop over the control streams that use the created data set, you are all set to rerun the models as needed.

## Stamp objects for traceability
The last couple of functions that will be introduced here are used for
tracing datasets to data creation scripts, including time stamps and
other information you want to include with the data set. 

```{r}
pk <- NMstamp(pk,script="vignettes/DataCreate.Rmd")
NMinfo(pk)
```
The `script` argument is recognized by `NMstamp`, but you can add anything to this. Say you want to keep descriptive note too:

```{r}
pk <- NMstamp(pk,script="vignettes/DataCreate.Rmd",Description="A PK dataset used for examples.")
NMinfo(pk)
```

These are very simple functions. But they are simple to use as well,
and hopefully they will help you avoid sitting with a data set trying
to guess which script generated it so you can do a modification or
understand how something was done.

When using `NMwriteData`, you don't have to call `NMstamp`
explicitly. Just pass the `script` argument to `NMwriteData` and
`NMstamp` will be applied automatically.


## Subject-level variables
In a proper PK/PD analysis we need to explore the data at multiple
variability levels. What we looked at above is at dosing and sampling
level (one row per dosing or sampling event). `NMdata` provides very
useful functions to extract information at other levels of
variability. Before extracting the subject-level information we will
add an individual exposure measure to the dataset. We will use the
empirical Bayes' estimate of the individual maximum
concentration. This is derived as the maximum prediction across the
sample times - it may be better to simulate the model at a richer time
scale to get better precision. Again, we make use of `data.table` but
feel free to use what you prefer. `findCovs` - like other `NMdata`
functions - will work on `data.frame`'s, `tibble`'s and other
`data.frame`-like classes 
(classes that inherit from `data.frame`).

```{r}
res1 <- NMscanData(system.file("examples/nonmem/xgxr001.lst", package="NMdata"),
                   col.row="ROW",merge.by.row=TRUE,quiet=TRUE,as.fun="data.table")
res1$trtact <- reorder(res1$trtact,res1$DOSE)
## with data.table, create a new column representing ID-level Cmax
res1[,Cmax:=max(IPRED),by=.(ID)]
## findCovs picks the columns that do not vary within cols.id. One row
## per value of cols.id.
res1.id <- findCovs(res1,by="ID")
dim(res1.id)
ggplot(res1.id,aes(WEIGHTB,Cmax/DOSE,colour=trtact))+
    geom_point()+
    labs(x="Bodyweight at baseline (kg)")
```

`by` can be longer than one. Include a study column, a treatment column to
support cross-over, or a model
column if you are comparing model fits (see below how to do this very easily).

If your model includes occasion variability, you probably also want to
look at 

```{r}
## we have no occasion variability in this data
## res1.id.occ <- findCovs(res1,by=c("ID","OCC"))
```

The simplest use of `findCovs` is to get variables that are constant
across the whole dataset:


```{r}
findCovs(res1)
```

Let's take a look at what is in the
`res1.id` generated above. It is a mix of variables that vary at subject level and variables
that are constant across the full dataset. 

```{r}
dim(res1.id)
head(res1.id,2)
```


`findCovs` has a counterpart in `findVars` which finds variables that
do vary within constant values of optional `by` columns. To get only the ones that
vary within the dataset (i.e. they are truely subject-level
variables and not say a study number if we only have one study), we can do 

```{r}
res1.id2 <- findVars(res1.id)
dim(res1.id2)
head(res1.id2,2)
```

`findVars` supports the `by` argument too. So you can use
`findVars(res1,by="ID")` to find variables that are non-constant
within (at least one value of) ID. `by` can be of arbitrary length.

Of course, we most often know what covariates or other subject-level
variables to look at, and we would not search for them after running a
nonmem model. But in the situation where you are looking at someone
else's work or you are doing a meta analysis across models where the
data has been coded slightly differently, these simple tools can be
very useful. Or if you like writing generic wrapper functions, you may
find this very handy. Also, remember that if a variable is returned by
these functions, you know that they fullfill the variability
requirement. We know that variables in `res1.id` above do not vary
within `ID`, meaning that no `ID` has more than one value of the
variable. `NA` counts as any other value, so if a value is returned
for subject in `res1.id`, this subject does not have `NA`'s in `res1`.

