---
title: "Data Preparation Tools"
output:  rmarkdown::html_vignette
Suggests: markdown
VignetteBuilder: knitr
vignette: >
  %\VignetteIndexEntry{Data Preparation Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{ae}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
 ,fig.width=7)

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

```{r,eval=TRUE,include=FALSE}
## library(devtools)
## load_all("C:/Users/delff/working_copies/NMdata")
```

```{r,setup,include=F}
library(NMdata)
NMdataConf(check.time=FALSE)
## NMdataConf(as.fun="data.table")
library(data.table)
library(ggplot2)
theme_set(theme_bw()+theme(legend.position="bottom"))

```
Built `r Sys.Date()` using NMdata `r packageVersion("NMdata")`.

This vignette is still under development. Please make sure to see
latest version available
[here](https://philipdelff.github.io/NMdata/).

## Introduction
Getting data ready for modeling is a crucial and often underestimated
task. Mistakes during the process of combining data sets, defining
time variables etc. can lead to difficulties during modeling, need for
revisiting data set preparation, and in worst case wasted time working
with an erroneos data set. Avoiding those mistakes by integrating
checks into the data preparation process is a key element in an efficient
and reliable data preparation work flow.

Furthermore, NONMEM has a number of restrictions on the format
of the input data, and problems with the data set is a common
reason for NONMEM not to behave as expected. When this happens,
debugging can be time-consuming. `NMdata` includes some simple
functions to prevent these situations.

This vignette uses `data.table` syntax for the little bit of data
manipulation performed. However, you don't need to use data.table _at
all_ to use these or any tool in `NMdata`. The data set is a
`data.table`:

```{r}
pk <- readRDS(file=system.file("examples/data/xgxr2.rds",package="NMdata"))
class(pk)
```

If you are not familiar with `data.table`, you can still keep reading this
vignette and learn what `NMdata` can do. `data.table`
is a powerful enhancement to the `data.frame` class, and the syntax is a little different from `data.frame`. The few places where this affects the examples
provided here, explanations will be given.

<!-- Because the data is a `data.table`  -->

<!-- The reason why the -->
<!-- functions return data.table objects is that -->

<!-- If this was a `data.frame` (and not a `data.table`), the `NMdata` functions would keep returning objects -->
<!-- of class `data.frame`. -->

## Data assembly


### Compare presence and class of columns across data sets
When stacking (`rbind`) and merging, it is most often necessary to check
if two or more data sets are compatible for the
operation. `compareCols` compares columns across two or more data sets.

```{r,include=FALSE}
pk.reduced <- copy(pk)
pk.reduced <- pk.reduced[1:(.N%/%2)]
pk.reduced[,CYCLE:=NULL]
pk.reduced[,AMT:=as.character(AMT)]
```

To illustrate the output of `compareCols`, a slightly modified
version of the `pk` dataset has been created. One column (`CYCLE`) has
been removed, and `AMT` has been recoded to character. `compareCols`
tells us about exactly these two differences: 

```{r,eval=TRUE}
compareCols(pk,pk.reduced)
```

Before merging or stacking, we may want to recode `AMT` in one of the
datasets to get the class we need, and decide what to do about the
`CYCLE` column which is missing in one of the datasets (add information or fill with
`NA`?).


### Rename columns based on contents
The model estimation step is heavily dependent (and in NONMEM almost entirely based) on numeric data values. 
The source data will often contain character variables, i.e. columns with non-numeric data values.

If the column names reflect whether the values are numeric, double-checking can be avoided. `renameByContents` renames columns if a function of their contents returns `TRUE`.


```{r,include=FALSE}
pktmp <- copy(pk)
pktmp[,TRTACT:=NULL]
```

```{r}
pk.renamed <- renameByContents(data=pktmp,fun.test=NMisNumeric,fun.rename = tolower,
                               invert.test = TRUE)
```

We make use of the function `NMisNumeric` which tests if NONMEM can
interpret the contents as numeric. If say the subject ID is of
character class, it can be valid to NONMEM. Subject ID `"1039"` will
be a numeric in NONMEM, `"1-039"` will not. `NMisNumeric` will return
`TRUE` if and only if all elements are either missing or interpretable
as numeric. We invert the condition (`invert.test=TRUE`), and _the
names_ of the columns that NONMEM cannot interpret as numeric become
lowercase. We use `compareCols` to illustrate that three columns were
renamed:

```{r,eval=TRUE}
compareCols(pktmp,pk.renamed)
```

We can now easily see that if we wish to include the information
contained in `eventu`, `pktmp`, and `pk.renamed`, we have to modify or
translate their contents first.

### Automated checks of merge results 
Merges are a very common source of data creation bugs. As simple as
they may seem, merges can leave you with an unexpected number of
rows, some repeated and/or some omitted. Often, we can impose restrictions
on the merge operation that allows for automated validation of the
results. 

Imagine the very common example that we have a longitudinal
PK data set (called `pk`), and we want to add subject-level covariates
from a secondary data set (`dt.cov`). We want to merge by `ID`, and
all we can allow to happen is columns to be added to `pk` from `dt.cov`. If rows
disappear or get repeated, or if columns get renamed, it's unintended
and should return an error.

```{r,include=FALSE}
## dt.cov <- pk[,.(ID=unique(ID)[1:10])]
## dt.cov[,COV:=sample(1:5,size=10,replace=TRUE)]
dt.cov <- pk[,.(ID=unique(ID))]
dt.cov[,COV:=sample(1:5,size=.N,replace=TRUE)]
dt.cov <- dt.cov[c(1,1:(.N-1))]
```

We merge the two data sets and the check of the dimensions raises no alarm - the number of rows is unchanged from `pk` to `pk2`, and one of two columns in `dt.cov` was added.
```{r}
pk2 <- merge(pk,dt.cov,by="ID")
dims(pk,dt.cov,pk2)
```

What we didn't realize is that we now have twice as many rows for subject 31.
```{r}
pk[ID==31,.N]
pk2[ID==31,.N]
```

Using `mergeCheck`, we get an error. This is because `mergeCheck` compares the actual rows going in and out of the merge and not just the dimensions:

```{r}
try(mergeCheck(pk,dt.cov,by="ID"))
```

Notice that `mergeCheck` tells us for which values of `ID` the input
and output differ so we can quickly look into the data sets and make a
decision how we want to handle this. In this case we discard the
covariate value for subject 31 and use `all.x=TRUE` argument to get
`NA` for subjects 31 and 180:

```{r}
dt.cov2 <- dt.cov[ID!=31]
pk2.check <- mergeCheck(pk,dt.cov2,by="ID",all.x=TRUE)
```

To ensure the consistency of rows before and after the merge, you
could use `merge(...,all.x=TRUE)` and then check dimensions before and
after (yes, both `all.x=TRUE` and the dimension check are necessary). This is not needed if you use `mergeCheck`.

`mergeCheck` does not try to reimplement merging. Under the hood, the
merging is performed by `data.table::merge.data.table` to which most
arguments are passed. What `mergeCheck` does is to add the checks that
the results are consistent with the criteria outlined
above. `data.table::merge.data.table` is generally very fast, and even
if there is a bit of extra calculations in `mergeCheck`, it should
never be slow.

Another problem  the programmer may not realize during a merge is
when column names are shared across `x1` and `x2` (in addition to
columns that are being merged by). This will silently create column
names like `col.x` and `col.y` in the output. `mergeCheck` will by
default give a warning if that happens (can be modified using the
`fun.commoncols` argument). Also, there is an optional argument to
tell mergeCheck how many columns are expected to be added by the
merge, and `mergeCheck` will fail if another number of columns are
added. This can be useful for programming.


The row order is by default maintained by `mergeCheck`. Apart from
this, there is only one difference from the behavior of the
`merge.data.frame` function syntax, being that the `by` argument must
always be supplied to `mergeCheck`. Default behavior of
`merge.data.frame` is to merge by all common column names, but for
coding transparency, this is intentionally not allowed by
`mergeCheck`.

In summary, `mergeCheck` verifies that the rows that result from the
merge are the exact same as in one of the existing datasets, only
columns added from the second input dataset. You may think that this
will limit your merges, and that you need merges for inner and outer
joins etc. You are exactly right - `mergeCheck` is not intended for
those merges and does not support them. When that is said, the kind of
merges that are supported by `mergeCheck` are indeed very common. All
merges in the `NMdata` package are performed with `mergeCheck`.


## Exclusion flags
It is good practice not to discard records from a dataset but to flag
them and omit them in model estimation. When reporting the analysis,
we also need to account for how many data records were discarded due
to which criteria. A couple of functons in `NMdata` help you do this
in a way that is easy to integrate with NONMEM. 

The implementation in `NMdata` is based on sequentially checking
exclusion conditions. This means we can summarize how many records and
subjects were excluded from the analysis due to the different
criteria. The information is represented in one numerical column for
NONMEM, and one (value-to-value corresponding) character column for
the rest of us in the resulting data.

### Assign and count flags
For use in NONMEM, the easiest is that inclusion/exclusion is
determined by a single column in data - we call that column `FLAG`
here, but any column name can be used. `FLAG` obviously draws on
information from other columns such as `TIME`, `DV`, and many others,
depending on your dataset and your way of working.


The function that applies inclusion/excluasion rules is called
`flagsAssign`, and it takes a dataset and a data.frame with rules as
arguments.

```{r,include=F}
pk <- readRDS(file=system.file("examples/data/xgxr2.rds", package="NMdata"))
pk[,`:=`(FLAG=NULL,flag=NULL)]
```
```{r}
dt.flags <- fread(text="FLAG,flag,condition
10,Below LLOQ,BLQ==1
100,Negative time,TIME<0")

pk <- flagsAssign(pk,tab.flags=dt.flags,subset.data="EVID==0")
pk <- flagsAssign(pk,subset.data="EVID==1",flagc.0="Dosing")
```

`fread` is used to create a data.table (like `read.csv` to create a
data.frame) for readability, one line for each row in the data.table created.
`flagsAssign` applies the conditions sequentially and by decreasing 
value of `FLAG`. `FLAG=0` means that the observation is included in
the analysis. You can use any expression that can be evaluated within
the data.frame. In this case, `BLQ` has to exist in `pk`.

Finally, flags are assigned to `EVID==1` rows. Here, no flag table is
used. This means that all `EVID==1` rows will get `FLAG=0`. 

In NONMEM, you can include `IGNORE=(FLAG.NE.0)` in `$DATA` or `$INFILE`.

Again, the omission will be attributed to the first condition
matched. Default is to apply the conditions by the order of decreasing
numerical flag value. Use `flags.increasing=TRUE` if you prefer the
opposite. However, what cannot be modified is that 0 is the numerical
value for rows that are not matched by any conditions.


What rows to omit from a dataset can vary from one analysis to
another. Hence, the aim with the chosen design is that the inclusion
criteria can be changed and applied to overwrite an existing
inclusion/exclusion selection. For another analysis we want to include
the observations below LLOQ. We have two options. Either we simply
change the `IGNORE` statement given above to `IGNORE=(FLAG.LT.10)`, or
you create a different exclusion flag for that one. If you prefer to
create a new set of exclusion flags, just use new names for the
numerical and the character flag columns so you don't overwrite the
old ones. See help of `flagsAssign` and `flagsCount` for how.


### Summarize data exclusions
An overview of the number of observations disregarded due to the
different conditions is then obtained using `flagsCount`:

```{r}
tab.count <- flagsCount(data=pk[EVID==0],tab.flags=dt.flags)
print(tab.count)
```

`flagsCount` includes a `file` argument to save the the table right
away.


## Finalize data format and write to file
Once the dataset is in place, `NMdata` provides a few useful functions
to ensure the formatting of the written data is compatible with
NONMEM. These functions include checks that NONMEM will be able to
interpret the data as intended, and more features are under
development in this area.

### Automated ordering of columns
The order of columns in NONMEM is important for two reasons. One is
that a character in a variable read into NONMEM will make the run
fail. The other is that there are restrictions on the number of
variables you can read into NONMEM, depending on the
version. `NMorderColumns` tries to put the used columns first, and
other or maybe even unusable columns in the back of the dataset. It
does so by a mix of recognition of column names and analysis of the
column contents. 

Columns that cannot be converted to numeric are put in the back, while
column bearing standard NONMEM variable names like `ID`, `TIME`, `EVID` etc.
will be pulled up front. You can of course add column names to
prioritize to front (`first`) or back (`last`). See `?NMorderColumns` for
more options.

```{r}
pk <- NMorderColumns(pk)
```

We may want to add `MDV` and rerun `NMorderColumns`.


### Writing data to files
For the final step of writing the dataset, `NMwriteData` is
provided. Most importantly, it writes a csv file with appropriate options for NONMEM to read it as well as possible. It can also write an rds for
R with equal contents (or RData if you prefer), but with the rds including all information
(such as factor levels) which cannot be saved in csv. If you should use `NMscanData` to read NONMEM results,  this information can be used automatically. It also provides a proposal for text to include in the
`$INPUT` and `$DATA` sections of the NONMEM control
streams. 

```{r}
NMwriteData(pk)
```
Notice, `NMwriteData` detected the exclusion flag and suggests to include it in `$DATA`.

If a file name had been provided, the data would have been written,
and the path to the data file would have been included in the message
written back to the user. There are several arguments that will affect
the proposed text for the NONMEM run, see `?NMwriteData`.

### Updating NONMEM control streams to read new data file
I may be the only one, but sometimes during the modeling stage, I want
to go back and change or add something to the data creation step. Then
once I have written a new data file, my NONMEM `$INPUT` sections no
longer match the data file. In `NMwriteData` you can use a `last`
argument to get columns pushed towards the back so the NONMEM runs
should still work, but maybe you need the column in your nonmem runs,
and so you have no way around updating the control streams. And that
can be quite a lot of control streams.

`NMdata` has a couple of functions to extract and write sections to
NONMEM control streams called `NMreadSection` and `NMwriteSection`. We
are not going into detail with what these functions can do, but let's
stick to the example above. We can do

```{r,eval=FALSE}
NMwriteSection("run001.mod","INPUT","$INPUT ROW ID TIME EVID CMT AMT DV FLAG STUDY BLQ CYCLE DOSE FLAG2 NOMTIME PART PROFDAY PROFTIME WEIGHTB eff0")
```

But in fact, we can go a step further and take the information straight from NMwriteData

```{r}
text.nm <- NMwriteData(pk)
```

NMwriteData invisibly returns a list of sections (`$INPUT` and `$DATA`). `NMwriteSection` can use these directly. So to write only the `$INPUT` section to `run001.mod`, we do the following. Please notice the single brackets in `text.nm["INPUT"]` which mean that we still send a list to `NMwriteSection`.

```{r,eval=FALSE}
NMwriteSection("run001.mod",list.sections=text.nm["INPUT"])
```

If you run this in a loop over the control streams that use the created data set, you are all set to rerun the models as needed.

## Stamp objects for traceability
The last couple of functions that will be introduced here are used for
tracing datasets to data creation scripts, including time stamps and
other information you want to include with the data set. 

```{r}
pk <- NMstamp(pk,script="vignettes/DataCreate.Rmd")
NMinfo(pk)
```
The `script` argument is recognized by `NMstamp`, but you can add anything to this. Say you want to keep descriptive note too:

```{r}
pk <- NMstamp(pk,script="vignettes/DataCreate.Rmd",Description="A PK dataset used for examples.")
NMinfo(pk)
```

These are very simple functions. But they are simple to use as well,
and hopefully they will help you avoid sitting with a data set trying
to guess which script generated it so you can do a modification or
understand how something was done.

When using `NMwriteData`, you don't have to call `NMstamp`
explicitly. Just pass the `script` argument to `NMwriteData` and
`NMstamp` will be applied automatically.


