---
title: "NMdata Cheatsheet"
output:
rmarkdown::html:
    toc: true
Suggests: markdown
VignetteBuilder: knitr
vignette: >
  %\VignetteIndexEntry{NMdata Cheatsheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{ae}
---

```{css,echo=FALSE}
.main-container {
	max-width: unset;
	margin: 50px;
}

.column-container{
	/* background-color: hotpink */
	display:flex;
	flex-direction:row;
	justify-content:space-between;

}
.column-container__column{
	/* background-color:green; */
	width:calc(25% - 24px);
}

.Code {
	/* background-color: red; */
	/* background-color: #a3edd5; */
	background-color: #c4ffec;
}

```

<!-- Todo: -->
<!-- merge.by.row=F -->
<!-- better explanation for flagsAssign -->


```{r, include = FALSE}
##knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(
                      collapse = TRUE
                     ,comment = "#>"
                     ,fig.width=7
                     ,cache=FALSE
                     ,class.source="Code"
                  )
library(data.table)
library(NMdata)
## library(devtools)
## load_all()

NMdataConf(as.fun="data.table"
          ,check.time=F)

library(ggplot2)
theme_set(theme_bw()+theme(legend.position="bottom"))

## this change data.table syntax. I think we can do without.
## knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)

pk <- readRDS(file=system.file("examples/data/xgxr2.rds",package="NMdata"))
## pk[,trtact:=NULL]
covs <- unique(pk[,.(ID,WEIGHTB)])
pk[,WEIGHTB:=NULL]
set.seed(1)
covs2 <- covs[,.(ID,race=sample(c("caucasian","black"),size=.N,replace=T))]

```

<!-- The NMdata logo -->
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("apple-touch-icon-180x180.png"),
               alt = 'logo', 
               style = 'position:absolute; top:15px; right:70px; padding:0px; width:150px')
```


:::: {style="display: flex;" class="column-container"}
::: {class="column-container__column"}

Built `r Sys.Date()` using NMdata `r packageVersion("NMdata")`.

Please make sure to see latest version available
[here](https://philipdelff.github.io/NMdata/). 

This cheat sheet is
intented to provide an overview and remind of command names. Please
refer to other vignettes for more details on specefic topics and
individual manual pages for details on the functions.

The order of the contents loosely follows a workflow example (except for
the configuration part). The steps can be applied in any order and
independently from each other.


### Get started
```{r,eval=FALSE}
install.packages("NMdata")
library(NMdata)
```


### Data preparation
We have read some source data files and need to combine them into one data set for NONMEM. Key steps are stacking data sets (like doses, samples, and simulation records) and adding additional information such as covariates. We often use `rbind` and merge or join operations for these steps. NMdata helps explore how to do these steps and ensure that merge/join results are as expected.

<!-- `dims` - compare dimensions of multiple data sets -->
<!-- ```{r} -->
<!-- ## In these examples we use a PK dataset and a set of covariates -->
<!-- dims(pk,covs) -->
<!-- ``` -->

`compareCols` - Compare presence and classes of columns across data sets before merging or stacking. 
```{r}
compareCols(covs,covs2)
```
Use the `cols.wanted` argument for the overview to especially focus on the columns you need in your final data set. 


`mergeCheck(x,y,...)` - Merges data and only accept results if all that happened was that columns from `y` were added to `x`. Row order of `x` is retained. Arguments are passed to data.table which does the actual merge. This completely automates the necessary checks when say merging covariates onto data.

```{r}
pk2 <- mergeCheck(pk,covs2,by="ID")
```
We did not get an error from `mergeCheck` so we know that the rows in `pk2` are exactly identical to those in `pk`, except the addition of a column called `cov2`. If rows duplicate or disappear `mergeCheck` does a good job telling you where in data to address the issues. 

`renameByContents` - Keep track of what columns are compatible with NONMEM by renaming columns accordingly. `NMisNumeric` evaluates whether NONMEM can interpret contents as numeric (different from `is.numeric`):
```{r,eval=FALSE}
## Example 1: Append an "N" to columns that NONMEM _can_ read (as numeric)
pk <- renameByContents(data=pk,
                       fun.test = NMisNumeric,
                       fun.rename = function(x)paste0(x,"N"))
## Example 2: lowercase names of columns that NONMEM _cannot_ read
pk <- renameByContents(data=pk,
                       fun.test = NMisNumeric,
                       fun.rename = tolower,
                       invert.test = TRUE)
```


`flagsAssign` - Sequentially assign exclusion flags to a dataset based
on a set of user-specified exclusion criteria.

`flagsCount` - Create an overview of number of retained and discarded datapoints.

Example with only two exclusion flags applied to samples. If time is negative, we assign exclusion flag `FLAG=100`. If (time is non-negative and) `BLQ==1` we assign `FLAG=10`. If none of these conditions are met, `FLAG=0`, and the row is to be included in the analysis. (`fread` is just for row-wise readability.)

```{r,include=FALSE}
pk[,(cc(FLAG,flag)):=NULL]
```

```{r}
dt.flags <- fread(text="FLAG,flag,condition
10,Below LLOQ,BLQ==1
100,Negative time,TIME<0")
pk <- flagsAssign(pk,tab.flags=dt.flags,subset.data="EVID==0")
pk <- flagsAssign(pk,subset.data="EVID==1",flagc.0="Dosing")
flagsCount(pk[EVID==0],tab.flags=dt.flags)[,.( flag, N.left, Nobs.left, N.discard, Nobs.discard)]
```
You may also want to apply a couple of exclusion criteria to dose records (for missing time, zero or missing amounts?) by modifying the steps above and applying to `EVID==1`.

`NMorderColumns` - Standardize column order. Columns that can be read
by NONMEM are prioritized. Row identifier and standard column names have special priorities.
```{r,include=FALSE}
pk <- NMorderColumns(pk)
```

`NMcheckData` - Extensive data checks for NONMEM compatibility and common issues. Should be run before saving data but see the "Debuging..." section for example on output.

:::

::: {class="column-container__column"}

### Write data and update NONMEM control streams accordingly
`NMwriteData` - Write data ensuring compatibility with NONMEM. By defaults saves both a csv (for NONMEM) and an rds (for R, retaining factor levels etc). Text for optional use in `$INPUT` and `$DATA` NONMEM sections is returned. `script` and `args.stamp` are optional arguments, see "Traceability" section for their purpose.

```{r}
text.nm <- NMwriteData(pk,file="derived/pkdata.csv",script="NMdata-cheat.Rmd",args.stamp=list(Description="PK data for the NMdata Cheatsheet"))
```

```{r,include=FALSE}
## writing version 2 rds so we don't need to depend on R 3.5
NMwriteData(pk,file="derived/pkdata.csv",script="NMdata-cheat.Rmd",args.stamp=list(Description="PK data for the NMdata Cheatsheet"),args.rds=list(version=2))
```

`NMwriteSection` - Replace sections of a NONMEM control stream. Can use the text generated by `NMwriteData` to update NONMEM runs to match the newly generated input data. Update INPUT section (and not DATA) for all control streams in directory "nonmem" which file names start with "run1" and end in ".mod" (say "run101.mod" to "run199.mod"):
```{r,eval=FALSE}
NMwriteSection(dir="nonmem",
               file.pattern="run1.*\\.mod",
               list.sections=text.nm["INPUT"])
```
`NMwriteSection` has the argument `data.file` to further limit the scope of files to update based on what data file the control streams use. It only makes sense to use the auto-generated text for control streams that use this data set.

The text for NONMEM is generated by `NMgenText`. Use that to generate alternative `$INPUT` sections (e.g. for models that use other columns as dependent variables) without saving data again. You can tailor the generation of the text to copy `(DV=CONC)`, drop `(COL=DROP)`, rename (`DV` instead of `CONC`) and more.

### Debugging input data (and control stream)
`NMcheckData` can check a `data.frame`. However, it can also be run on a path to a control stream, in which case it provides a full check of how data is read by NONMEM and then checks the data _as read by NONMEM_. It checks column names in INPUT section against data and then runs a full check of the data set as read by NONMEM (according to column names in $INPUT and ACCEPT/IGNORE statements in $DATA). We suppress the default print to terminal (`quiet=T`) and provide selected parts of the results here.

```{r}
res.debug <- NMcheckData(file="nonmem/run201.mod",quiet=T)
## we will only show some of what is available here
names(res.debug)
## Meta data on input data file:
res.debug$tables
```
In this model we forgot to update the control stream INPUT section after adding a column to data ("off" means that INPUT text can be reorganized to match data file better):
```{r}
## Comparison of variable naming:
res.debug$input.colnames[c(1:2)]
res.debug$input.colnames[c(9:12)]
```
We have some findings on the data set too. But since `res.debug$input.colnames` tells us we are reading the data incorrectly, we have to address that before interpreting findings on the data. 
```{r}
res.debug$NMcheckData$summary
```
If you are preparing a data set, run `NMcheckData` directly on the data (using the `data` argument) insted of on a control stream.

:::
::: {class="column-container__column"}

### Retrieve NONMEM results
`NMscanData` - Automatically find NONMEM input and output tables and organize data. By default, available column names are taken from the NONMEM control stream. Additional column names (columns not read by NONMEM) are taken from input data file.
```{r,include=F}
##NMscanData <- function(x)NMdata::NMscanData(file.path(system.file(paste0("examples/nonmem/",x), package="NMdata")))
## res1 <- NMscanData(system.file("examples/nonmem/xgxr001.lst", package="NMdata"))
```
```{r}
res1 <- NMscanData("nonmem/run101.lst")
```
```{r,include=F}
## rm(NMscanData)
```

The following plot serves to illustrate that the obtained data set combines output tables (`PRED` is from a $TABLE statement) with input data (exclusion flags are represented as character variables). Moreover, the "below LLOQ" samples are included in the result even though they were not in the analysis (excluded using `IGNORE` in control stream, recovered in `NMscanData` using `recover.rows=TRUE`)
```{r}
## Recover rows that were not read by NONMEM (due to ACCEPT/IGNORE)
res2 <- NMscanData("nonmem/run101.lst",recover.rows=TRUE)
```
```{r}
library(ggplot2)
res2.plot <- subset(res2,ID==135&EVID==0)
ggplot(res2.plot,aes(TIME))+
    geom_point(aes(y=DV,colour=flag))+
    geom_line(aes(y=PRED))+
    labs(y="Concentration (unit)",subtitle=unique(res2$model))
```

Read the messages from `NMwriteData` and `NMscanData` carefully and notice that an rds file was written and read. This bypasses the loss of information caused by writing and reading csv, and so we have kept factor levels from the input data we generated:

```{r}
levels(res1$trtact)
```
:::

::: {class="column-container__column"}

### Configuration

Use the many options in `NMdataConf` to tailor NMdata behaviour to your setup and preferences. Make NMdata functions return data.tables or tibbles:
```{r,eval=FALSE}
NMdataConf(as.fun=tibble::as_tibble)
NMdataConf(as.fun="data.table")
```

By default, NMdata functions will look for a unique row identifier in columns called `ROW`. If you call this column `REC`, do
```{r,eval=FALSE}
NMdataConf(col.row="REC")
```

By default, NMdata is configured to read files from PSN in which case the input control stream is needed to find the input data. Do this if you don't use PSN:
```{r,eval=FALSE}
NMdataConf(file.mod=identity)
```

Loosely speaking, `NMdataConf` changes default values of NMdata
function arguments. Many options can be configured this way so you
don't have to remember to type in those arguments every time you call
an NMdata funtion.


### Traceability 
`NMinfo` - Get metadata from an NMdata object. This will show where and when input data was created, when model was run, results of concistency checks, what tables were read, how they were combined and a complete list of data columns and their origin.


A list of the available elements:
```{r}
names(NMinfo(res1))
```
The information recorded during saving of the input data:
```{r}
NMinfo(res1,"dataCreate")
```
A full list of columns in all columns in output and input data is included. The source data file and the column number in the result (`COLNUM`) are listed.
```{r}
NMinfo(res1,"columns")[1:8]
```
We saw earlier that we got "30+2" columns back. We see that the additional two were added by NMscanData (`source`). `DV` was already included from another table so the redundant `DV` column is omitted.
```{r}
NMinfo(res1,"columns")[30:33]
```

### Additional functions to read (NONMEM) data files
`NMscanTables` - Find and read all output data tables based on a NONMEM control stream file. A list of tables is returned.

`NMreadTab` - Read an output table file from NONMEM based on path to output data file

`NMscanInput` - Read input data based on NONMEM control stream and optionally translate column names according to the `$INPUT` NONMEM section

`NMreadCsv` - Read input data formatted for NONMEM

::: 
::::
