---
title: "NMdata Cheatsheet"
output:
rmarkdown::html:
    toc: true
Suggests: markdown
VignetteBuilder: knitr
vignette: >
  %\VignetteIndexEntry{NMdata Cheatsheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{ae}
---
<style>
.main-container {
max-width: unset;
margin: 50px;
}

.column-container{
/* background-color: hotpink */
display:flex;
flex-direction:row;
justify-content:space-between;

}
.column-container__column{
/* background-color:green; */
width:calc(25% - 24px);
}
</style>

```{r, include = FALSE}
##knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(
                      collapse = TRUE
                     ,comment = "#>"
                     ,fig.width=7
                     ,cache=FALSE
                  )
library(data.table)
## library(NMdata)
library(devtools)
load_all()

NMdataConf(as.fun="data.table"
           ,check.time=F)

## this change data.table syntax. I think we can do without.
## knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)

pk <- readRDS(file=system.file("examples/data/xgxr2.rds",package="NMdata"))
pk[,trtact:=NULL]
covs <- unique(pk[,.(ID,WEIGHTB)])
pk[,WEIGHTB:=NULL]
set.seed(1)
covs2 <- covs[,.(ID,cov2=sample(c("caucasian","black"),size=.N,replace=T))]

```

<!-- The NMdata logo -->
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(## file.path(R.home("pkgdown"), "favicon", "apple-touch-icon-152x152.png")),
                                ## system.file("pkgdown/favicon/apple-touch-icon-152x152.png",package="NMdata")),
                                "apple-touch-icon-180x180.png"),
               alt = 'logo', 
               style = 'position:absolute; top:50px; right:50px; padding:0px; width:150px')
```

Built `r Sys.Date()` using NMdata `r packageVersion("NMdata")`.

Please make sure to see latest version available
[here](https://philipdelff.github.io/NMdata/). 

:::: {style="display: flex;" class="column-container"}
::: {class="column-container__column"}

This cheat sheet is
intented to provide an overview and remind of command names. Please
refer to other vignettes for more details on specefic topics and
individual manual pages for details on the functions.

### Basics
Get started
```{r,eval=FALSE}
install.packages("NMdata")
library(NMdata)
```

Use the many options in `NMdataConf` to tailor NMdata behaviour to your setup and preferences. Make NMdata functions return data.tables or tibble's:
```{r,eval=FALSE}
NMdataConf(as.fun=tibble::as_tibble)
NMdataConf(as.fun="data.table")
```

By default, NMdata is configured to read files from PSN in which case the input control stream is needed to find the input data. Do this if you don't use PSN:
```{r,eval=FALSE}
NMdataConf(file.mod=identity)
```

Loosely speaking, `NMdataConf` changes default values of NMdata function arguments.

### Data set preparation


<!-- `dims` - compare dimensions of multiple data sets -->
<!-- ```{r} -->
<!-- ## In these examples we use a PK dataset and a set of covariates -->
<!-- dims(pk,covs) -->
<!-- ``` -->

`compareCols` - Compare presence and classes of columns across data sets before merging or stacking.
```{r}
compareCols(covs,covs2)
```



:::

::: {class="column-container__column"}
### Data set preparation cont'd
`renameByContents` - Keep track of what columns are compatible with Nonmem by renaming those that aren't. Rename all columns that Nonmem cannot interpret as numeric to lowercase (see `NMisNumeric` in Programming section):
```{r,echo=F}
pk <- renameByContents(data=pk,
                       fun.test = NMisNumeric,
                       fun.rename = tolower,
                       invert.test = TRUE)
## Or append a "C" to the same column names
pk <- renameByContents(data=pk,
                       fun.test = NMisNumeric,
                       fun.rename = function(x)paste0(x,"C"),
                       invert.test = TRUE)
```

`mergeCheck(x1,x2)` - Merge and only accept results if all that happened was that columns were added to `x1`.

`flagsAssign` - Assign exclusion flags to a dataset based on specified table

`flagsCount` - Create an overview of number of retained and discarded datapoints

`NMorderColumns` - Standardize column order in Nonmem input data


`NMcheckData` - Check data for Nonmem compatibility and common issues.

`NMstamp` - Add metadata for documentation and traceability of datasets. Information does not affect the data itself but can be stored together with it and accessed with `NMinfo`.

`NMwriteData` - Write data ensuring compatibility with NONMEM. Text for optional use in `$INPUT` `$DATA` NONMEM sections are returned. `NMwriteSection` can use this information to update NONMEM runs to match the newly generaed input data.

### Read from and edit Nonmem control streams
`NMreadSection` - Extract sections of Nonmem control streams

`NMwriteSection` - Replace sections of a nonmem control stream

### Debugging Nonmem and input data
`NMcheckColnames` - Compare `$INPUT` in control stream to column names in input data. 
```{r,include=F}
NMcheckColnames <- function(x)NMdata::NMcheckColnames(file.path(system.file(paste0("examples/nonmem/",x), package="NMdata")))
```
```{r}
NMcheckColnames("xgxr001.lst")[1:3,]
```
```{r,include=F}
rm(NMcheckColnames)
```

:::

::: {class="column-container__column"}

### Retrieve Nonmem results
`NMscanData` - Automatically find Nonmem input and output tables and organize data
```{r,include=F}
NMscanData <- function(x)NMdata::NMscanData(file.path(system.file(paste0("examples/nonmem/",x), package="NMdata")))
## res1 <- NMscanData(system.file("examples/nonmem/xgxr001.lst", package="NMdata"))
```
```{r}
res1.dt <- NMscanData("xgxr003.lst")
class(res1.dt)
NMdataConf(as.fun=tibble::as_tibble,quiet=TRUE)
res1.tbl <- NMscanData("xgxr003.lst")
class(res1.tbl)
dims(res1.dt,res1.tbl)
```
```{r,include=F}
rm(NMscanData)
```

`NMscanTables` - Find and read all output data tables based on a NONMEM control stream file. A list of tables is returned.

`NMreadTab` - Read an output table file from NONMEM based on path to output data file

`NMscanInput` - Read input data based on NONMEM control stream and optionally translate column names according to the `$INPUT` NONMEM section

`NMreadCsv` - Read input data formatted for nonmem

`NMinfo` - Get metadata from an NMdata object. This will show where and when input data was created, when model was run, results of concistency checks, what tables were read, how they were combined and a complete list of data columns and their origin.

:::


::: {class="column-container__column"}

### Traceability 
NMinfo

```{r}
names(NMinfo(res1.dt))
NMinfo(res1.dt,"datafile")
NMinfo(res1.dt,"input.colnames")
```

:::
::::
