---
title: "NMdata FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMdata FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{ae}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
 ,fig.width=7)

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```


Built `r Sys.Date()` using NMdata `r packageVersion("NMdata")`.


DISCLAIMER: This vignette is under development. 

## I want to use a tool from NMdata, but how can I avoid working with data.table?
Easy. Every function in `NMdata` has an argument called `fun.as`. You
can provide a function in this argument which will be applied to data
before returnin it.

```{r,eval=F}
NMscanData(...,as.fun=as.data.frame)
```

You can also get a tibble if you work with `tidyverse` tools: 

```{r,eval=F}
NMscanData(...,as.fun=tibble::as_tibble)
```

If you want to change the behaviour generally for the session and omit
the `as.fun` argument, you can do this by this option:

```{r,eval=F}
options(NMdata.as.fun=tibble::as_tibble)
```

And all `NMdata` functions (that return data) should now return tibbles.

