---
title: "Parameter Tables"
output:
rmarkdown::html_vignette:
    toc: true
Suggests: markdown
VignetteBuilder: knitr
vignette: >
  %\VignetteIndexEntry{02ParameterTables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{ae}
---

```{r,include = FALSE}
## library(tidyvpc)
library(NMdata)
load_all("~/wdirs/NMdata")
NMdataConf(as.fun="data.table")

##knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(
                      collapse = TRUE
                     ,comment = "#>"
                     ,fig.width=7
                     ,cache=FALSE
                  )

## NMdataConf(dir.psn="/opt/psn")

library(data.table)
library(knitr)
library(scales)
library(NMcalc)
```

Parameter tables are a basic and essential look into a model estimate. Due to Nonmem's flexible model definition approach, getting an informative parameter table can however be an involved process. 

* Nonmem parameters like THETA(1), OMEGA(2,2), SIGMA(1,1) etc. must be connected with physiologically interpretable parameters such as "Clearance" or "Between-subject variability on central volume" to be informative. 

* Physiologically interpretable parameters may be transformations of Nonmem parameters.

A manual or even semi-automated approach to establishing these relationships are manually involved to quality check. 

`NMdata` offers two distinct approaches to connecting Nonmem parameters to physiologially interpretable ones.

* Flexible, user-defined interpretation of delimited comments in the parameter sections ($THETA, $OMEGA, and $SIGMA) (`NMreadParText()`). This means, if you added comments in whatever consistent way in these sections, `NMreadParText()` is likely to be able to get it into a formatted `data.frame`.

* Automatic translation of Nonmem code related to direct use of the parameters (`NMrelate()`). This may be a pretty good approach trying to report a Nonmem model without structured labeling in the parameter sections. If you keep a  coding standard, it may allow you to skip manual parameter translation altogether. And lastly, it may be a powerful QC tool in any case.

## Reading Nonmem parameter estimates
Before getting into how to translate the Nonmem parameters, it is worth mentioning how to even just read the parameter estimates with `NMdata`. A lot of the relevant information is stored in the `.ext` file and can be read using `NMreadExt()`.

```{r}
## file.mod <- system.file("examples/nonmem/xgxr001.mod",package="NMdata")
## file.mod <- system.file("examples/nonmem/xgxr132.mod",package="NMdata")
file.mod <- system.file("examples/nonmem/xgxr133.mod",package="NMdata")

##NMreadExt(file.mod,return="objv")
pars.ext <- NMreadExt(file.mod)

```

This does provide a basic parameter table.

```{r}

pars.ext[,.(par.name,FIX,value,se)]

```

`NMreadExt()` does return more than that. These are all the columns available:

```{r}
colnames(pars.ext)
```
where `est` is a deprecated copy of the `value` column.

NMreadExt can return other information than the parameter estimates. This vignette is about creating parameter tables so for now, let this be a teaser for what else `NMreadExt()` provides (which is all the info in `.ext` files):

```{r}
names(NMreadExt(file.mod,return="all"))
```

## Adding physiologically interpretable labels

The above parameter table is not useful to many, maybe not even to the person who wrote the model. Let

```{r}
NMreadSection(file.mod,section=c("THETA"))
NMreadSection(file.mod,section=c("OMEGA"))
NMreadSection(file.mod,section=c("SIGMA"))

```

```{r}
partab <- NMreadParsText(file.mod,format="%init;%idx:%symbol;%label[%unit];%trans",
                         format.sigma="%init;%symbol-%label")
partab
```
And we can format a parameter table with labels merging with the ext file:

```{r}
partab2 <- mergeCheck(pars.ext,
           partab,
           by="parameter",
           all.x=TRUE,
           fun.commoncols="drop.y",
           quiet=TRUE
           )
partab2[,value.trans:=value]
partab2[trans=="log",value.trans:=exp(value)]
kable(partab2[,.(par.name,symbol,FIX,label,unit,Estimate=value.trans,RSE=se/value)])

```

The left join we did with `mergeCheck` includes all parameters from the ext file. In this case we want to skip all the off-diagonals in $OMEGA and $SIGMA because they are not estimated. Let's just look at the onese that were estimated. In many cases some fixed parameters can be relevant to include in a parameter table, and then you need to modify this step. 

```{r}
partab2[FIX==0,.(par.name,symbol,FIX,label,unit,Estimate=NMcalc::signif2(value.trans,3),RSE=scales::percent(se/value,accuracy=.1))] |>
    kable()
```

We need to explain that RSE's were calculated on the log scale for the
transformed parameters. We could also want to add confidence intervals
based on the estimated standard error on the log scale. But we are
getting close.

### Add shrinkage
```{r}
shk <- NMreadShk(file.mod)

partab3 <- mergeCheck(partab2,shk[,.(par.name,Pval=signif2(Pval,2),etabar,EtaShkSD=percent(EtaShkSD/100))],by="par.name",all.x=TRUE)
partab3[FIX==0,.(par.name,symbol,FIX,label,unit,Estimate=NMcalc::signif2(value.trans,3),RSE=scales::percent(se/value,accuracy=.1),Pval,etabar,EtaShkSD)] |>
    kable()
```

### Correlation estimates

```{r}

Partab4 <- NMdata::addOmegaCorr(partab2)
kable(Partab4)
```

### Editing information in the parameter table
The problem with the otherwise nicely looking table generated above is
that we spotted a typo in one label. A "t" is missing in
"Intercompartmental". How are we going to fix that? The model was
finalized so we don't want to edit the control stream. For this
specific problem we could fix it with a simple text replacement. But
such patchwork can become messy with a large model. While the method
of providing all the details in control stream may look like a great
solution at first, it has problems

A practical issue is the tedious process keeping all the candidate
models up to date as the information for the parameter tables
improves. The parameter table work is often part of a model reporting
step following model development. Since the control streams are
created in the model development step, the parameter table fine tuning
is difficult to integrate. Briefly, we need a way to update the
parameter tables after finalizing the control stream.

```{r,include=FALSE,eval=FALSE}
### dont do this - csv has been edited
tab.csv <- partab[,.(symbol,label,unit)]
fwrite(tab.csv,file="derived/partab_offline.csv")
```

```{r}
tab.offline <- read.csv("derived/partab_offline.csv")
mergeCheck(partab2[FIX==0],tab.offline,by="symbol",fun.commoncols="drop.x")
```


### Identifying the parameter names from code
In the last example, all we used from the control stream was the `symbol`, i.e. an abbreviated parameter name. The rest was merged in from a table maintained in a csv file.


<!-- off-diags missing. Merge info from ext? -->
```{r}
load_all("~/wdirs/NMdata")
NMdataConf(as.fun="data.table")

labs.auto <- NMrelate(file.mod)
labs.auto[,symbol:=label]
labs.auto[par.type=="OMEGA",symbol:=paste0("BSV.",symbol)]
kable(labs.auto)

mergeCheck(labs.auto[,.(par.name,symbol)],tab.offline,by="symbol",all.x=T)
```

## More models
Any of the code above can be run on multiple models at once. `NMdata` functions generally return tables including a `model` column for this purpose. Let's get a quick overview of a few models.

```{r}
files <- system.file(file.path("examples/nonmem",c("xgxr132.mod","xgxr133.mod")),package="NMdata")
labs.auto <- NMrelate(files)
labs.auto[,symbol:=label]
labs.auto[par.type=="OMEGA",symbol:=paste0("BSV.",symbol)]

NMreadExt(files,return="pars")[FIX==0]|>
    mergeCheck(labs.auto[,.(model,par.name,symbol)],by=c("model","par.name"),all.x=T) |> 
    mergeCheck(tab.offline,by="symbol",all.x=T) |>
    dcast(par.name+symbol+label+unit~model,value.var="value")


```
