::: {class="column-container__column"}

### Programming

`cc` - create character vectors without typing quotations unless special characters are used
```{r}
cc(A , B, "A-B")
```

Imagine you want to plot random effects against covariates. `findCovs` and `findVars` are fast functions that extract the columns that vary a given level of variability (between subjects, between occasions).

`findCovs` - Extract columns that do not vary within values of other columns
```{r}
## columns that don't vary at all. Could be a study number or an ETA that is not used.
(model.level <- findCovs(pk))
## columns that don't vary within subjects (by can be of arbitrary length)
id.level <- findCovs(pk,by="ID")
```

`findVars` - Extract columns that _vary_ within values of other columns
```{r,eval=FALSE}
## in id.level, some do not vary between ID's so in fact they are model-level. Discard those
actual.id.level <- findVars(id.level,by="ID")
## find occasion level variability - columns that do vary between ID but do not vary with ID+OCC.
occ.level <- findCovs(findVars(pk,"ID"),c("ID","OCC"))
```


`NMextractDataFile` - Get the file path to the input data file used in a NONMEM control stream

`NMisNumeric` - Test whether a column or the individual values in a column/vector are compatible with NONMEM (interpretable as numeric)

<!-- `tmpcol` -->
<!-- `egdt` -->
 

:::
